---
title: "Afrobarometer"
author: "Juliana Gómez Consuegra"
date: "`r Sys.Date()`"
output: pdf_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(haven) #load spss file
library(dplyr) #data manipulation
library(foreign)
library(ggplot2)
library(tidyr)
```


```{r load_data, include=FALSE}
#with haven
# afro_spss <- read_sav('data/afrobarometer_2019/r7_merged_data_34ctry.release.sav')

#with foreign
afro_spss <-read.spss("data/afrobarometer_2019/r7_merged_data_34ctry.release.sav",  
                   to.data.frame = TRUE) 

head(afro_spss)
```

```{r format_names, include=FALSE}
# change all . in afro_spss column names to _
names(afro_spss) <- gsub("\\.", "_", names(afro_spss))
```

```{r subset_df, include=FALSE}
country_list = c('Tanzania', 'Kenya', 'Uganda', 'Mozambique')
questions_columns = c('RESPNO','COUNTRY','LOCATION_LEVEL_1','Q38D', 'Q38E', 'Q38F', 'Q77A', 'Q77B', 'Q77C', 'Q77D', 'Q78B', 'Q86A', 'Q94', 'Q95C')

afro_subset <- afro_spss %>% 
  select(all_of(questions_columns)) %>% 
  filter(COUNTRY %in% country_list)
```

# Labels for our variables of interest

|QUESTIONS|LABEL|LEVEL|
|---|---|---|
|Q38D|Missing  |-1|   
|Q38E|Strongly Disagree |1|
|Q38F|Disagree |2 |
|Q77B|Neither Agree Nor Disagree   |3 | 
|Q77C|Agree  |4 |   
|Q77D|Strongly Agree |5|
||Refused   |8 |  
||Don’t Know |9|
|---------|
|Q77A|Missing  |-1    |     
||Strongly Disagree |1|
||Disagree |2 |
||Neither Agree Nor Disagree   |3 |                   
||Agree  |4 |             
||Strongly Agree |5|
||Refused   |8 |             
||Don’t Know |9|
||Not asked in the country |99|
|---------|
|Q78B|Missing  |-1    | 
||Never justified |1| 
||Sometimes justified |2 |  
||Always justified |3| 
||Refused  |8 |        
||Don't know | 9 |
|---------|
|Q86A| Missing  | -1 |      
||Never |0 |
||Once or twice |1| 
||Several times  |2|  
||Many times  |3|      
||Refused |8|
||Don't know |9|
|---------|
|Q94|Missing |-1| 
||No, not looking |0|     
||No, looking |1|  
||Yes, part time |2|  
||Yes, full time |3| 
||Refused |8|     
||Don’t know |9|
|---------|
|Q95C|Missing |-1|
||You make the decisions yourself |1| 
||Your spouse makes the decisions |2| 
||You make the decisions jointly with your spouse |3|
||You make the decisions jointly with other family members |4| 
||Other family members make the decisions without you |5|
||None of the above / Some other response |6| 
||Not applicable, no earnings |7|
||Refused |8|
||Don’t know |9| 

```{r check_for_nulls, include=FALSE}
#check for nulls in afro_subset. No NULLS
print(afro_subset %>%
  summarise(count = sum(is.na(.))))
```


```{r levels, include=FALSE}
# check levels for all variables
for (i in 3:12) {
  print(names(afro_subset)[i])
  print(levels(afro_subset[[i]]))
}
```


```{r format_variables, include=FALSE}
# change datatype in vars
test <- afro_subset

#format variables
change_minusone_to_nine <- function(x) {
  x <- x %>% 
    factor(levels = c("Missing", "Strongly Disagree", "Disagree",
                      "Neither Agree Nor Disagree", "Agree", "Strongly Agree",
                      "Refused", "Don’t Know"),
           labels = c('-1', '1', '2', '3', '4', '5', '8', '9'))
  return(x)
}


for_one_to_nine_vars <- c('Q38D', 'Q38E', 'Q38F', 'Q77C', 'Q77B', 'Q77D')
for (i in for_one_to_nine_vars) {
  test[[i]] <- change_minusone_to_nine(afro_subset[[i]])
}

# format Q77A
test$Q77A <- test$Q77A %>% 
  factor(levels = c("Missing", "Strongly Disagree", "Disagree",
                    "Neither Agree Nor Disagree", "Agree", "Strongly Agree",
                    "Refused", "Don’t Know", "Not asked in the country"),
         labels = c('-1', '1', '2', '3', '4', '5', '8', '9', '99'))


# format Q78B
test$Q78B <- test$Q78B %>% 
  factor(levels = c("Missing","Never justified","Sometimes justified","Always justified",                    "Refused","Don't know"),
         labels = c('-1', '1', '2', '3', '8', '9'))

#format Q86A
test$Q86A <- test$Q86A %>% 
  factor(levels = c("Missing","Never","Once or twice", "Several times",
                    "Many times", "Refused","Don't know"),
         labels = c('-1', '0', '1', '2', '3', '8', '9'))

#format Q94
test$Q94 <- test$Q94 %>% 
  factor(levels = c("Missing","No, not looking", "No, looking","Yes, part time",
                    "Yes, full time",  "Refused","Don’t know"),
         labels = c('-1', '0', '1', '2', '3', '8', '9'))

#format Q95C
test$Q95C <- test$Q95C %>% 
  factor(levels = c('Missing', 'You make the decisions yourself', 'Your spouse makes the decisions', 'You make the decisions jointly with your spouse', 'You make the decisions jointly with other family members', 'Other family members make the decisions without you', 'None of the above / Some other response', 'Not applicable, no earnings', 'Refused', "Don’t know"),
         labels = c('-1', '1', '2', '3', '4', '5', '6', '7', '8', '9'))

#chage datatypes to numeric for all Q
for (i in 4:14) {
  test[[i]] <- as.numeric(test[[i]])
}

# typecast to relevant types
test$LOCATION_LEVEL_1 <- as.factor(test$LOCATION_LEVEL_1)

#check for nas (still 8 NAs)
print(test %>%
   summarise(count = sum(is.na(.))))

```


```{r summary_statistics_unstandarized, include=FALSE}
# summary statistics for all variables, group by location_level_1 Drop RESPNO


summary_stats <- test %>% 
  select(!RESPNO)%>% 
  group_by(COUNTRY,LOCATION_LEVEL_1) %>% 
  summarise_all(list(mean = mean, median = median, sd=sd)) %>% 
  # summarise_all(funs(mean(., na.rm = TRUE), sd(., na.rm = TRUE), sum(!is.na(.)))) %>% 
  ungroup()

summary_stats
```


```{r summary_statistics_standarized, include=FALSE}
# standardize the data. Make sure to account for divisions by zero
standardize <- function(x) {
  # Calculate mean and standard deviation
  mean_x <- mean(x, na.rm = TRUE)
  sd_x <- sd(x, na.rm = TRUE)
  
  # Check for zero standard deviation
  if (sd_x == 0) {
    # Add a small epsilon value to avoid division by zero
    epsilon <- 1e-10  # You can adjust this value if needed
    return((x - mean_x) / (sd_x + epsilon))
  } else {
    # Standard calculation if standard deviation is not zero
    return((x - mean_x) / sd_x)
  }
}

# change missing data to NA and standardize
test_standarized <- test %>% 
  select(!RESPNO)%>% 
  mutate_all(~replace(., . == -1, NA))%>%
  group_by(COUNTRY,LOCATION_LEVEL_1) %>%
  mutate_all(standardize) %>%
  ungroup()


summary_stats_standarized <- test_standarized %>%
  group_by(COUNTRY,LOCATION_LEVEL_1) %>%
  summarise_all(list(mean = mean, median = median, sd=sd)) %>%
  ungroup()

#check for NaN in test_standarized
print(test_standarized %>%
  summarise(count = sum(is.na(.))))

```

```{r boxplots_standarized, include=TRUE, fig.width=10, fig.height=10}
# boxplots for all variables in test_standarized. Use ggplot and panels
boxplot_standardized <- test_standarized %>% 
  gather(key = "variable", value = "value", -COUNTRY, -LOCATION_LEVEL_1) %>% 
  ggplot(aes(x = variable, y = value)) +
  theme_bw()+
  geom_boxplot() +
  facet_wrap(~COUNTRY, scales = "free") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Boxplots for all variables in test_standarized", x = "Variables", y = "Values")

print(boxplot_standardized)
```


```{r}
#export summary_stats_standarized to csv
write.csv(summary_stats_standarized, "data/afrobarometer_2019/summary_stats_standarized.csv")
```

                                   

  
                                                 
                                                       

