---
title: "Afrobarometer 2022"
author: "Juliana Gómez Consuegra"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(haven) #load spss file
library(dplyr) #data manipulation
library(foreign)
library(ggplot2)
library(tidyr)
```


```{r load_data, include=FALSE}
#with foreign
afro_2022 <-read.spss("data/afrobarometer_2022/afrobarometer_release-dataset_merge-34ctry_r8_en_2023-03-01.sav",  
                   to.data.frame = TRUE) 

head(afro_2022)
```

```{r format_names, include=FALSE}
# change all . in afro_2022 column names to _
names(afro_2022) <- gsub("\\.", "_", names(afro_2022))
```

Questions in 2022 are different from 2019: 

|---|---|---|
|Variable label|2019|2022|
|Respondent number|RESPNO|RESPNO|
|Country|COUNTRY|COUNTRY|
|Location level|LOCATION_LEVEL_1|REGION|
|Men have more right to job|Q38D|X|
|Women have equal right to land|Q38E|X|
|Better if woman takes care of household|Q38F|X|
|Girls and boys have equal chance at education|Q77A|X|
|Women and men have equal chance to earn income|Q77B|X|
|Women and men have equal chance of paying job|Q77C|X|
|Women and men have equal chance to own/inherit land|Q77D|X|
|Justified for men to beat their wives|Q78B|X|
|Experienced discrimination based on gender|Q86A|X|
|Employment status|Q94 |Q95A|
|Who decides how money is used|Q95C|X|


We will use the following questions related to gender in 2022 (where one of the options is= 28 Gender issues/women's rights) for our analysis. 

|---|---|---|
|Question number|Variable label|
|Q48PT1|Most important problems – 1st response|
|Q48PT2|Most important problems – 2nd response|
|Q48PT3|Most important problems – 3rd response|


```{r subset_df, include=FALSE}
country_list = c('Tanzania', 'Kenya', 'Uganda', 'Mozambique')
questions_columns = c('RESPNO','COUNTRY','REGION','Q48PT1','Q48PT2', 'Q48PT3','Q95A')

afro_subset <- afro_2022 %>% 
  select(all_of(questions_columns)) %>% 
  filter(COUNTRY %in% country_list)
```


```{r check_for_nulls, include=FALSE}
#check for nulls in afro_subset. No NULLS
print(afro_subset %>%
  summarise(count = sum(is.na(.))))
```


Change labels to levels: 

Pt1: 0-33, 9995, 9998, 9999, -1
Pt2: 1-33, 9995, 9996
Pt3: 1-33, 9996

```{r to_lower, include=FALSE}
# change values in Q48PT1, Q48PT2, Q48PT3 to lower case
afro_subset$Q48PT1 <- tolower(afro_subset$Q48PT1)
afro_subset$Q48PT2 <- tolower(afro_subset$Q48PT2)
afro_subset$Q48PT3 <- tolower(afro_subset$Q48PT3)

# trim ws in values for Q48PT1, Q48PT2, Q48PT3
afro_subset$Q48PT1 <- trimws(afro_subset$Q48PT1)
afro_subset$Q48PT2 <- trimws(afro_subset$Q48PT2)
afro_subset$Q48PT3 <- trimws(afro_subset$Q48PT3)

# Define a function to replace spaces around backslashes
replace_specific_around_all_slashes <- function(x) {
  # Use gsub() with a capturing group and backreference
  gsub("(?<=\\w)(\\s*)(\\/)(?=\\w)", "\\1/", x, perl = TRUE)  # Escape slash with \\/
}


afro_subset$Q48PT1 <- replace_specific_around_all_slashes(afro_subset$Q48PT1)
afro_subset$Q48PT2 <- replace_specific_around_all_slashes(afro_subset$Q48PT2)
afro_subset$Q48PT3 <- replace_specific_around_all_slashes(afro_subset$Q48PT3)

```



```{r levels, include=FALSE}
# check levels for all variables
afro_subset[,4:7] <- lapply(afro_subset[,4:7], factor)


for (i in 4:7) {
  print("--------------------------------")
  print(names(afro_subset)[i])
  print(levels(afro_subset[[i]]))
}

```





```{r format_variables, include=FALSE}
afro_subset <- afro_subset %>% mutate(Q48PT1=recode(Q48PT1,
                            "agricultural marketing" = 32,                        
                            "aids" = 21,
                            "civil war" = 31,
                            "climate change" = 33,
                            "communications" = 12,
                            "corruption" = 24,
                            "crime and security" = 23,
                            "democracy/political rights" = 29,
                            "discrimination/ inequality" = 27,
                            "don’t know" = 9999,
                            "drought" = 9,
                            "education" = 14,
                            "electricity" = 16,
                            "farming/agriculture" = 7,
                            "food shortage/famine" = 8,
                            "gender issues / women’s rights" = 28,
                            "health" = 20,
                            "housing" = 15,
                            "infrastructure / roads" = 13,
                            "land" = 10,
                            "loans / credit" = 6,
                            "management of the economy" = 1,
                            "nothing/ no problems" = 0,
                            "orphans/street children/homeless children" = 18,
                            "other responses" = 9995,
                            "political instability/political divisions/ ethnic tensions" = 26,
                            "political violence" = 25,
                            "poverty/destitution" = 4,
                            "rates and taxes" = 5,
                            "refused" = 9998,
                            "services (other)" = 19,
                            "sickness / disease" = 22,
                            "transportation" = 11,
                            "unemployment" = 3,
                            "wages, incomes and salaries" = 2,
                            "war (international)" = 30,
                            "water supply" = 17)) %>% 
mutate(Q48PT2=recode(Q48PT2,
                            "agricultural marketing" = 32,                        
                            "aids" = 21,
                            "civil war" = 31,
                            "climate change" = 33,
                            "communications" = 12,
                            "corruption" = 24,
                            "crime and security" = 23,
                            "democracy/political rights" = 29,
                            "discrimination/ inequality" = 27,
                            "drought" = 9,
                            "education" = 14,
                            "electricity" = 16,
                            "farming/agriculture" = 7,
                            "food shortage/famine" = 8,
                            "gender issues / women’s rights" = 28,
                            "health" = 20,
                            "housing" = 15,
                            "infrastructure / roads" = 13,
                            "land" = 10,
                            "loans / credit" = 6,
                            "management of the economy" = 1,
                            "no further reply" = 9996,
                            "orphans/street children/homeless children" = 18,
                            "other responses" = 9995,
                            "political instability/political divisions/ ethnic tensions" = 26,
                            "political violence" = 25,
                            "poverty/destitution" = 4,
                            "rates and taxes" = 5,
                            "refused"= 9998,
                            "services (other)" = 19,
                            "sickness / disease" = 22,
                            "transportation" = 11,
                            "unemployment" = 3,
                            "wages, incomes and salaries" = 2,
                            "war (international)" = 30,
                            "water supply" = 17))%>%
mutate(Q48PT3=recode(Q48PT3,
                            "agricultural marketing" = 32,                        
                            "aids" = 21,
                            "civil war" = 31,
                            "climate change" = 33,
                            "communications" = 12,
                            "corruption" = 24,
                            "crime and security" = 23,
                            "democracy/political rights" = 29,
                            "discrimination/ inequality" = 27,
                            "drought" = 9,
                            "education" = 14,
                            "electricity" = 16,
                            "farming/agriculture" = 7,
                            "food shortage/famine" = 8,
                            "gender issues / women’s rights" = 28,
                            "health" = 20,
                            "housing" = 15,
                            "infrastructure / roads" = 13,
                            "land" = 10,
                            "loans / credit" = 6,
                            "management of the economy" = 1,
                            "no further reply" = 9996,
                            "orphans/street children/homeless children" = 18,
                            "other responses" = 9995,
                            "political instability/political divisions/ ethnic tensions" = 26,
                            "political violence" = 25,
                            "poverty/destitution" = 4,
                            "rates and taxes" = 5,
                            "services (other)" = 19,
                            "sickness / disease" = 22,
                            "transportation" = 11,
                            "unemployment" = 3,
                            "wages, incomes and salaries" = 2,
                            "war (international)" = 30,
                            "water supply" = 17))%>% 
mutate(Q95A=recode(Q95A,
                            "No (not looking)" = 0,
                            "No (looking)" = 1,
                            "Yes, part time" = 2,
                            "Yes, full time" = 3,
                            "Refused" = 8,
                            "Don’t know" = 9,
                            "Missing" = -1))

```

```{r gender_issues, include=FALSE}
# calculate frequencies of answers for each level, for each question

afro_subset$gender_issues <- ifelse(afro_subset$Q48PT1 == 28 | afro_subset$Q48PT2 == 28 | afro_subset$Q48PT3 == 28, 1, 0)

afro_subset_freq <- afro_subset %>% 
  select(REGION,COUNTRY,gender_issues) %>% 
  group_by(COUNTRY, REGION,gender_issues) %>% 
  summarise(n = n()) %>% 
  mutate(prop = n/sum(n))




#plot bloxplot for each question
afro_subset_freq %>%
filter(gender_issues ==1)%>%
ggplot(aes(x = COUNTRY, y = prop, fill = COUNTRY)) +
geom_boxplot() +
theme_bw()+
scale_fill_brewer(palette = "Set3")+
labs(x= "Country", y = "Proportion of responses", title = "Proportion of respondents who mentioned gender issues as the most important issue in the country")



```








```{r}
#export afro_subset_freq to csv
write.csv(afro_subset_freq, "data/afrobarometer_2022/summary_stats_standarized.csv")
```

              